---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
library(geochem)
library(dplyr)
```

# geochem

<!-- badges: start -->
<!-- badges: end -->

**geochem** is an R package for wrangling, analysing, and visualising
geochemical data. It provides functions for element conversion, desurveying and projecting drillhole data, classification diagrams, spider plots, and publication-quality laser maps.

## Installation

The development version can be installed from
[GitHub](https://github.com/muhohl/geochem) with:

``` r
# install.packages("devtools")
devtools::install_github("muhohl/geochem")
```

## Built-in datasets

The package includes three synthetic datasets that cover the most common
geochemical workflows:

| Dataset | Description |
|---|---|
| `whole_rock_data` | 50 whole-rock analyses (Basalt / Andesite / Dacite / Rhyolite) with major oxides in wt % and trace elements in ppm |
| `laser_map_data` | 2 500-point 50 × 50 laser map with 8 element channels |
| `detection_limit_data` | 40 PGE assay rows with below-detection values stored as `"<X"` strings |

```{r datasets}
dplyr::glimpse(whole_rock_data)
```

---

## Examples

### Imputing below-detection-limit values

`random_number_imputer()` replaces `"<X"` strings with a uniform random
value drawn from `[0, X]`.

```{r bdl}
dl_imputed <- random_number_imputer(
  data = detection_limit_data,
  columns = 2:5,
  split_symbol = "<"
)

head(dl_imputed)
```

---

### Converting oxides to elements

`oxides_to_elements()` converts wt % oxide columns for common elements (Na₂O, MgO, Al₂O₃, …)
to wt % element using stoichiometric mass ratios. Column matching is
case-insensitive.

```{r oxides}
elements_pct <- oxides_to_elements(whole_rock_data)
dplyr::glimpse(elements_pct)
```

---

### Removing outliers with `filter_quantiles()`

`filter_quantiles()` replaces values beyond a quantile threshold with `NA`.
Below we remove the top 5 % of Fe and Cu from the laser map.

```{r quantiles}
laser_clean <- filter_quantiles(
  data = laser_map_data,
  .cols = c("Fe_ppm", "Cu_ppm"),
  probs = 0.95,
  upper_or_lower = "upper"
)

summary(laser_clean[c("Fe_ppm", "Cu_ppm")])
```

---

### TAS classification diagram

`geom_tas_diagram()` returns ggplot2 layers for the Total Alkali Silica
classification. Map SiO₂ to x and Na₂O + K₂O to y.

```{r tas}
library(ggplot2)

whole_rock_data |>
  dplyr::mutate(TotalAlkali = Na2O + K2O) |>
  ggplot(aes(x = SiO2, y = TotalAlkali)) +
  geom_tas_diagram() +
  geom_point(aes(colour = rock_type), size = 2.5, alpha = 0.85) +
  scale_colour_brewer(palette = "Set1", name = "Rock type") +
  labs(x = "SiO₂ (wt %)", y = "Na₂O + K₂O (wt %)") +
  theme_classic(base_size = 13)
```

---

### REE spider plot

`ggspider()` produces faceted spider diagrams with median lines and
interquartile ribbons. Pass `normalization = "mcdonough_sun_1995"` (or another
reference) to divide each element by its chondrite value before plotting; the
y-axis label updates to "REE/Chondrite" automatically.

```{r spider, fig.height = 5}
ree <- c("La", "Ce", "Nd", "Sm", "Eu", "Gd", "Yb", "Lu")

ggspider(
  data = whole_rock_data,
  elements = c("rock_type", ree),
  group = "rock_type",
  levels = ree,
  normalization = "mcdonough_sun_1995"
)
```

---

### Laser element maps

`laser_map2()` produces publication-quality elemental maps from laser-ablation
point data. The colour scale is chosen automatically based on the column name:
viridis for regular elements, a diverging gradient for ratios or PCA scores,
and Okabe–Ito for cluster maps. It uses `geom_raster()`, which means that the 
function can be applied to any 2-dimensional, equally spaced data set.
Ideally used with `patchwork::wrap_plots()` or `cowplot::plot_grid()` functions.

```{r laser, fig.height = 4}
maps <- laser_map2(
  data = laser_map_data,
  columns = c(3, 4, 7), # Fe_ppm, Cu_ppm, Au_ppb
  trans = "log",
  unit = ""
)

# Fe and Cu map
patchwork::wrap_plots(maps[1:2])
```

---

## Function overview

| Category | Functions |
|---|---|
| Data wrangling | `random_number_imputer()`, `oxides_to_elements()`, `niggli_numbers()`, `filter_quantiles()`, `fsp_molar()` |
| Classification diagrams | `geom_tas_diagram()`, `geom_magnetite_deposit()`, `geom_magnetite_origin()`, `ggfspmolar()` |
| Distribution & QQ plots | `ggcoloredqq()`, `ggspider()` |
| PCA | `pca_plot()`, `geom_pca_arrows()`, `eigen_value_plot()` |
| Laser maps | `laser_map2()`, `clipping_element()`, `raster_map()` (`laser_map()` deprecated) |
| Drillholes | `desurvey_holes()`, `drillhole_section()`, `raster_section()` |
| XMOD | `xmod_df_wrangler()`, `xmod_double_mapping()`, `xmod_samples_list()` |

For full documentation and worked examples see `vignette("introduction", package = "geochem")`.
