---
title: "Introduction to geochem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to geochem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Overview

**geochem** provides tools for wrangling, analysing, and visualising geochemical
data. This vignette walks through the main function categories using three
built-in synthetic datasets.

```{r load}
library(geochem)
library(dplyr)
library(ggplot2)
```

---

## Built-in datasets

The package ships with three synthetic datasets that mirror typical
laser-ICP-MS and whole-rock geochemistry workflows.

### `whole_rock_data`

Fifty whole-rock analyses across four volcanic rock types (Basalt, Andesite,
Dacite, Rhyolite). Columns include major oxides in wt % and trace elements
in ppm.

```{r wr-glimpse}
dplyr::glimpse(whole_rock_data)
```

### `laser_map_data`

A 50 × 50 point laser map (2 500 analyses) with x/y coordinates and eight
element channels in ppm / ppb. A synthetic ore zone is embedded near the
centre.

```{r lm-glimpse}
dplyr::glimpse(laser_map_data)
```

### `detection_limit_data`

Forty PGE assay results where a fraction of values fall below the instrument
detection limit, stored as strings such as `"<0.05"`.

```{r dl-glimpse}
head(detection_limit_data)
```

---

## 1 · Data preparation

### Imputing below-detection-limit values

`random_number_imputer()` replaces `"<X"` strings with a random value
uniformly drawn from `[0, X]`.

```{r random-imputer}
dl_imputed <- random_number_imputer(
  data         = detection_limit_data,
  columns      = 2:5,   # Au_ppb, Pd_ppb, Pt_ppb, Ag_ppb
  split_symbol = "<"
)

head(dl_imputed)
```

### Converting oxides to elements

`oxides_to_elements()` converts wt % oxide columns to wt % element using
stoichiometric mass ratios. Column names are matched case-insensitively
against standard oxide formulae (Na₂O, MgO, Al₂O₃, …).

```{r oxides-to-elements}
elements_pct <- oxides_to_elements(whole_rock_data)
dplyr::glimpse(elements_pct)
```

### Niggli numbers

`niggli_numbers()` normalises a major-oxide composition to Niggli parameters
following the "Niggli Modified" scheme of McDonald (2024). The four primary
numbers (`al`, `c`, `alk`, `fm`) always sum to 100. Both oxide wt % and
element wt % column names are accepted.

```{r niggli}
niggli <- niggli_numbers(whole_rock_data)
dplyr::glimpse(niggli)
```

### Removing outliers with `filter_quantiles()`

`filter_quantiles()` replaces values beyond a chosen quantile threshold with
`NA`, leaving the rest of the data frame intact.

```{r filter-quantiles}
laser_clean <- filter_quantiles(
  data           = laser_map_data,
  .cols          = c("Fe_ppm", "Cu_ppm"),
  probs          = 0.95,
  upper_or_lower = "upper"
)

summary(laser_clean[c("Fe_ppm", "Cu_ppm")])
```

### Feldspar molar ratios

`fsp_molar()` converts Na, K, and Al columns (oxide or element wt %) to
molar Na/Al and K/Al ratios. Column names containing `"O"` or `"o"` are
treated as oxides automatically; all others as element wt %.

```{r fsp-molar}
fsp <- whole_rock_data |>
  fsp_molar(na = "Na2O", k = "K2O", al = "Al2O3")

dplyr::select(fsp, Sample_ID, rock_type, na_al_molar, k_al_molar) |> head()
```

---

## 2 · Classification diagrams

### TAS diagram

`geom_tas_diagram()` returns ggplot2 layers for the Total Alkali Silica
classification. Map `SiO2` to x and `Na2O + K2O` to y.

```{r tas}
whole_rock_data |>
  dplyr::mutate(TotalAlkali = Na2O + K2O) |>
  ggplot2::ggplot(ggplot2::aes(x = SiO2, y = TotalAlkali)) +
  geom_tas_diagram() +
  ggplot2::geom_point(ggplot2::aes(colour = rock_type), size = 2.5, alpha = 0.85) +
  ggplot2::scale_colour_brewer(palette = "Set1", name = "Rock type") +
  ggplot2::labs(x = "SiO₂ (wt %)", y = "Na₂O + K₂O (wt %)") +
  ggplot2::theme_classic(base_size = 13)
```

### Feldspar molar diagram

`ggfspmolar()` plots the albite–orthoclase join as a molar ratio diagram.
Add `geom_point()` on top to display samples.

```{r fsp-plot}
fsp |>
  ggfspmolar() +
  ggplot2::geom_point(
    ggplot2::aes(colour = rock_type),
    size = 2
  ) +
  ggplot2::scale_colour_brewer(palette = "Set1", name = "Rock type") +
  ggplot2::theme_bw()
```

### Magnetite deposit-type diagram

`geom_magnetite_deposit()` draws the Ti+V vs Al+Mn discrimination fields
from Dupuis & Beaudoin (2011). Map Ti+V wt % to x and Al+Mn wt % to y on
log₁₀ scales.

```{r mgt-deposit, fig.height = 5}
ggplot2::ggplot() +
  geom_magnetite_deposit() +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::theme_bw()
```

### Magnetite origin diagram

`geom_magnetite_origin()` draws the Ni/Cr vs Ti boundary line from
Dare et al. (2014) separating magmatic from hydrothermal magnetite.
Both axes are on log₁₀ scales.

```{r mgt-origin, fig.height = 5}
ggplot2::ggplot() +
  geom_magnetite_origin() +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::theme_bw()
```

---

## 3 · Distribution plots

### REE spider plot

`ggspider()` produces faceted spider diagrams — one panel per group — with
median lines and interquartile ribbons. Pass a `normalization` reference to
divide by chondrite values before plotting.

```{r spider-plot, fig.height = 6}
ree <- c("La", "Ce", "Nd", "Sm", "Eu", "Gd", "Yb", "Lu")

ggspider(
  data      = whole_rock_data,
  elements  = c("rock_type", ree),
  group     = "rock_type",
  levels    = ree,
  normalization = "mcdonough_sun_1995"
)
```

### QQ plots with group colouring

`ggcoloredqq()` produces a quantile-quantile diagram for a single element,
with points coloured by a grouping variable. Because `geom_qq()` does not
support per-point colour mapping, theoretical quantiles are computed via
`ppoints()` and plotted with `geom_point()`.

```{r qq-plot}
ggcoloredqq(
  data  = whole_rock_data,
  element = "Cu",
  group = "rock_type"
) +
  ggplot2::scale_color_brewer(palette = "Set1", name = "Rock type") +
  ggplot2::theme_bw()
```

Use `lapply()` to produce plots for multiple elements at once:

```{r qq-multi, eval = FALSE}
lapply(c("Cu", "Ni", "Zn"), \(el) {
  ggcoloredqq(whole_rock_data, el, group = "rock_type") +
    ggplot2::theme_bw()
})
```

---

## 4 · Laser element maps

### Raster binning

`raster_map()` bins continuous x/y coordinates into an n × n regular grid,
useful for regularising irregular laser-ablation paths before plotting or
further analysis.

```{r raster-map}
raster_df <- raster_map(
  data = laser_map_data,
  x    = x,
  y    = y,
  n    = 25
)

dplyr::glimpse(raster_df)
```

### Element maps

`laser_map2()` produces publication-quality elemental raster maps. The colour
scale is chosen automatically from the column name: viridis for regular
elements, a diverging gradient for ratios or PCA scores, and Okabe–Ito for
cluster maps. Combine maps with `patchwork::wrap_plots()`.

```{r laser-map, fig.height = 4}
maps <- laser_map2(
  data    = laser_map_data,
  columns = c(3, 4, 7),   # Fe_ppm, Cu_ppm, Au_ppb
  trans   = "log",
  unit    = ""
)

patchwork::wrap_plots(maps[1:2])
```

---

## Session info

```{r session}
sessionInfo()
```
