---
title: "Introduction to geochem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to geochem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Overview

**geochem** provides tools for wrangling, analysing, and visualising geochemical
data, with particular emphasis on Laser-ICP-MS datasets. This vignette walks
through the main function categories using three built-in synthetic datasets.

```{r load}
library(geochem)
library(dplyr)
library(ggplot2)
```

---

## Built-in test datasets

The package ships with three synthetic datasets that mirror typical Laser-ICP-MS
and whole-rock geochemistry workflows.

### `whole_rock_data`

Fifty whole-rock analyses across four volcanic rock types (Basalt, Andesite,
Dacite, Rhyolite). Columns include major oxides in wt % and trace elements in ppm.

```{r wr-glimpse}
dplyr::glimpse(whole_rock_data)
```

### `laser_map_data`

A 50 × 50 point laser map (2 500 analyses) with x/y coordinates and eight
element channels in ppm / ppb. A synthetic ore zone is embedded near the centre.

```{r lm-glimpse}
dplyr::glimpse(laser_map_data)
```

### `detection_limit_data`

Forty PGE assay results where a fraction of values fall below the instrument
detection limit, stored as strings such as `"<0.05"`.

```{r dl-glimpse}
head(detection_limit_data)
```

---

## 1 · Imputing below-detection-limit values

`Random_Number_Imputer()` replaces `"<X"` strings with a random value uniformly
drawn from `[0, X]`.

```{r random-imputer}
dl_imputed <- Random_Number_Imputer(
  data = detection_limit_data,
  columns = 2:5, # Au_ppb, Pd_ppb, Pt_ppb, Ag_ppb
  split_symbol = "<"
)

head(dl_imputed)
```

---

## 2 · Converting oxides to elements

`oxides_to_elements()` converts wt % oxide columns to wt % element using
stoichiometric mass ratios. Column names are matched case-insensitively against
standard oxide formulae (Na2O, MgO, Al2O3, …).

```{r oxides-to-elements}
elements_pct <- oxides_to_elements(whole_rock_data)
dplyr::glimpse(elements_pct)
```

---

## 3 · Niggli numbers

`niggli_numbers()` normalises a major-oxide composition to the Niggli
parameters used in classical petrochemical classification.

```{r niggli}
niggli <- niggli_numbers(whole_rock_data)
dplyr::glimpse(niggli)
```

---

## 4 · Removing outliers with `filter_quantiles()`

`filter_quantiles()` replaces values beyond a chosen quantile threshold with
`NA`, leaving the rest of the dataframe intact. Here we remove the top 5 % of
Cu and Zn from the laser map data (quantile_position 19 = 95th percentile in
the default `probs = seq(0, 1, 0.05)` sequence).

```{r filter-quantiles}
laser_clean <- filter_quantiles(
  data = laser_map_data,
  .cols = list(3, 4), # Fe_ppm, Cu_ppm
  quantile_position = 19,
  upper_or_lower = "upper"
)

# NAs introduced at the top tail
summary(laser_clean[c("Fe_ppm", "Cu_ppm")])
```

---

## 5 · Spider (multi-element) plots

`spider_plot()` produces faceted spider diagrams — one panel per group — with
median lines and interquartile ribbons. The first element of the `elements`
vector is used as the grouping variable; the remainder are plotted on the x-axis.

```{r spider-plot, fig.height = 6}
trace_elements <- c("La", "Ce", "Nd", "Sm", "Eu", "Gd", "Yb", "Lu")
element_order <- trace_elements

spider_plot(
  data = whole_rock_data,
  elements = c("rock_type", trace_elements),
  group = "rock_type",
  levels = element_order,
  ribbon = TRUE
)
```

---

## 6 · Box plots by group

`box_plot()` produces one log-scaled boxplot per element, coloured by group, with
sample counts and mean values annotated.

```{r box-plot}
box_plot(
  data = whole_rock_data,
  elements = list(
    which(names(whole_rock_data) == "Cu"),
    which(names(whole_rock_data) == "Ni"),
    which(names(whole_rock_data) == "Zn")
  ),
  group = "rock_type"
)
```

---

## 7 · TAS classification diagram

`tas_diagram()` returns a list of ggplot2 layers that draw the Total Alkali
Silica fields. Map `SiO2` to x and `Na2O + K2O` to y, then add the template.

```{r tas, fig.height = 5}
whole_rock_data |>
  dplyr::mutate(TotalAlkali = Na2O + K2O) |>
  ggplot2::ggplot(ggplot2::aes(x = SiO2, y = TotalAlkali, colour = rock_type)) +
  geochem::tas_diagram() +
  ggplot2::geom_point(size = 2.5, alpha = 0.85) +
  ggplot2::scale_colour_brewer(palette = "Set1", name = "Rock type") +
  ggplot2::labs(x = "SiO₂ (wt %)", y = "Na₂O + K₂O (wt %)") +
  ggplot2::theme_classic(base_size = 13)
```

---

## 8 · Raster mapping

`raster_map()` bins continuous x/y coordinates into an n × n grid. This is
useful for regularising irregular laser-ablation paths before further analysis.

```{r raster-map}
raster_df <- raster_map(
  data = laser_map_data,
  x = x,
  y = y,
  n = 25 # 25×25 grid
)

dplyr::glimpse(raster_df)
```

---

## 9 · Laser element maps

`laser_map()` produces publication-quality elemental maps from laser-ablation
point data. It automatically selects a log or linear colour scale based on the
data range, calculates sensible legend breaks, and aligns the legend height to
the plot height.

```{r laser-map, fig.height = 4}
maps <- laser_map(
  data = laser_map_data,
  selected_elements = list(3, 4, 7), # Fe_ppm, Cu_ppm, Au_ppb
  Log_Trans = FALSE,
  unit_title = "(ppm / ppb)",
  fontsize = 11
)

# Display the Cu map
maps[[2]]
```

---

## Session info

```{r session}
sessionInfo()
```
